[**efg's R Notes:  Input/Output**](../index.html)

Reading Excel Files
===================

This page shows several examples of reading an Excel file into a data.frame.

Recall [**Sample.csv**](Sample.csv) file created using *writeLines*.  [View this csv on GitHub](https://github.com/EarlGlynn/earlglynn.github.io/blob/master/R/input-output/readcsv-writecsv/Sample.csv).

This CSV file was loaded into Excel .xlsx (Excel 2010) and .xls (Excel 2003) files.  Download [**Sample.xlsx**](Sample.xlsx) or [**Sample.xls**](Sample.xls).

Reading Excel files into R can be problematic.  Since I have not found one package that works with all Excel files, I show four approaches below. 

Find all the R code in this Gist, [R-IO-Excel.R](https://gist.github.com/EarlGlynn/8487321).

1. XLConnect Package: readWorksheetFromFile (uses Java)
-------------------------------------------------------

If using 64-bit Java, must use 64-bit R.

```{r}
# Avoid Java out-of-memory error for large Excel files.
# See p. 16, http://cran.r-project.org/web/packages/XLConnect/vignettes/XLConnect.pdf
options (java.parameters = "-Xmx1024m")
library(XLConnect)

d1 <- readWorksheetFromFile("Sample.xlsx", sheet=1)    # .xls or .xlsx
dim(d1)
str(d1)

d1
```

2. xlsx Package: read.xlsx or read.xlsx2 (uses Java)
-------------------------------------------------------

If using 64-bit Java, must use 64-bit R.

Read Excel file 2003 or earlier (.xls)
```{r}
library(xlsx)
d2 <- read.xlsx("Sample.xls", sheetIndex=1)   
dim(d2)
str(d2)
```

Read Excel file 2007 or later (.xlsx)
```{r}
d2a <- read.xlsx("Sample.xlsx", sheetIndex=1,
         stringsAsFactors=FALSE)
str(d2a)               
```

3. gdata Package: read.xls (uses PERL)
--------------------------------------

64-bit ActivePerl Community Edition works on Windows with either 32-bit or 64-bit R.  [http://www.activestate.com/activeperl](http://www.activestate.com/activeperl).

```{r}
library(gdata)

d3 <- read.xls("Sample.xls", sheet=1)    # .xls or .xlsx
dim(d3)
str(d3)

d3a <- read.xls("Sample.xlsx", sheet=1, as.is=TRUE)
dim(d3a)
str(d3a)

```

4. RODBC package:  sqlFetch or sqlQuery
---------------------------------------

Note:  "odbcConnectExcel is only usable with 32-bit Windows" with 32-bit R.

```
channel  <- odbcConnectExcel("Sample.xls")       # Excel 2003 and earlier
#channel <- odbcConnectExcel2007("Sample.xlsx")  # Excel 2007 and later
sqlTables(channel)

d4 <- sqlFetch(channel, "Sample")  # Worksheet name
#'data.frame':   4 obs. of  5 variables:
# $ Grade : Factor w/ 3 levels "A","B","F": 1 2 3 2
# $ Age   : num  23 56 NA 35
# $ Code  : Factor w/ 3 levels "ABC","DEF","XYZ": 1 2 NA 3
# $ Amount: num  48.98 2.72 0 3.14
# $ Start : POSIXct, format: "2014-01-01" "2014-01-02" NA "2014-01-05"

d4a <- sqlQuery(channel, paste("SELECT Code,Amount,Start,Age,Grade",
                               "FROM [Sample$]",
                               "WHERE Grade = 'B'"),
                as.is=TRUE)
dim(d4a)
#[1] 2 5

str(d4a)
#'data.frame':   2 obs. of  5 variables:
# $ Code  : chr  "DEF" "XYZ"
# $ Amount: num  2.72 3.14
# $ Start : chr  "2014-01-02 00:00:00" "2014-01-05 00:00:00"
# $ Age   : num  56 35
# $ Grade : chr  "B" "B"

close(channel)
```

My advice:  RODBC should only be used if you have an Excel file without holes, and with very regular numeric data. Without very regular data, RODBC can guess the wrong data types and cause other problems.

See [Madelaine's notes](http://research.stowers-institute.org/mcm/rmysql.html) for additional RODBC examples -- and some RMySQL examples.

*efg*, 2014-01-08
